$p$-adic precision

### 2014-01-30 padicprec.tex

Many usual softwares (\emph{e.g.} {\tt magma} and {\tt sage}) include 
primitives to handle $p$-adic numbers. Since $p$-adic numbers are by 
nature infinite, one needs to truncate them up to some finite level (we 
generally say \emph{finite precision}) in order to manipulate them on a 
computed. The usual technics to achieve this are inspired from case of 
real numbers: the computer works with ``truncated $p$-adic numbers'' 
represented as $x + O(p^n)$ and performs all basic operations (addition, 
product) on this representation. For instance, it computes: 
$$(x + O(p^n)) + (y + O(p^m)) = (x+y) + O(p^{\min(n,m)}).$$
However, in many important cases (\emph{e.g.} those discussed just 
above), this approach is too naive since it creates artificially 
important losses of precision (see \S\ref{ssec:stepbystep}). 
It is the so-called \emph{numerical instability} issue.

@@

Beyond sums and products (which can be treated as before), a basic 
operation involving polynomials is Euclidean division. Let us compute 
the differential of it. We start with two polynomials $A$ and $B$ with 
$B \neq 0$. The Euclidean division of $A$ by $B$ is written $A = BQ + R$ 
with $\deg R < \deg B$. Differentiating the above equality we find: 
$dA - dB \cdot Q = B \cdot dQ + dR$,
which implies that $dQ$ and $dR$ are respectively obtained as the 
quotient and the remainder of the Euclidean division of $dA - dB \cdot 
Q$ by $B$. This gives the differential. We note that the discussion 
above extends readily to convergent series (see also 
\cite{caruso-lubicz}).

@@

We now focus on the computation of Gr\"obner basis. Recall that any 
ideal $I \subset K[\XX]$ admits a unique reduced Gr\"obner basis. We can 
then consider the application mapping a family $(f_1, \ldots, f_s)$ of 
\emph{homogeneous}\footnote{This restriction will be convenient for us 
mainly because we are using the article \cite{Vaccon}, which deals only 
with homogeneous polynomials. It is however probably not essential.} 
polynomials of fixed degrees to the reduced Gr\"obner basis $(g_1, 
\ldots, g_t)$. It follows from Theorem 1.1 of \cite{Vaccon} that this 
application at $(f_1, \ldots, f_s)$ is continuous provided that:
\begin{itemize}
\item the sequence $(f_1, \ldots, f_s)$ is regular
\item for all $i \in \{1, \ldots, s\}$, the ideal generated by
$f_1, \ldots, f_i$ is wearly-$w$ where $w$ denotes the fixed monomial
order (\emph{cf} \cite{Vaccon} for more precisions).
\end{itemize}
A similar argument proves that it is actually differentiable at such
points. We are now going to compute the differential. For this we
remark that since the families $(f_1, \ldots, f_s)$ and $(g_1, \ldots,
g_t)$ generate the same ideal, we have a relation:
\begin{equation}
\label{eq:prodgrobner}
(g_1, \ldots, g_t) = (f_1, \ldots, f_s) \cdot A
\end{equation}
where $A$ is a $(s \times t)$ matrix with coefficients in $K[\XX]$.
Moreover following Vaccon's construction, we see that the entries of $A$ 
can be chosen in such a way that they define differentiable functions.
Differentiating Eq.~\eqref{eq:prodgrobner}, we get:
$$(d f_1, \ldots, d f_s) \cdot A = - (f_1, \ldots, f_s) \cdot dA + 
(d g_1, \ldots, d g_t).$$
Finally, using that we are working with \emph{reduced} Gr\"obner
basis, the above equality implies that $d f_i$ is the remainder in
the division of the $i$-th coefficient of the product
$(d f_1, \ldots, d f_s) \cdot A$ by the family $(f_1, \ldots, f_s)$.

@@

>>>>

Many usual softwares (\emph{e.g.} {\tt magma} and {\tt sage}) include 
primitives to handle $p$-adic numbers. Since $p$-adic numbers are by 
nature infinite, one needs to truncate them up to some finite level (we 
generally say \emph{finite precision}) in order to manipulate them on a 
computer. The usual technics to achieve this are inspired from case of 
real numbers: the computer works with ``truncated $p$-adic numbers'' 
represented as $x + O(p^n)$ and performs all basic operations (addition, 
product) on this representation. For instance, it computes: 
$$(x + O(p^n)) + (y + O(p^m)) = (x+y) + O(p^{\min(n,m)}).$$
However, in many important cases (\emph{e.g.} those discussed just 
above), this approach is too naive since it creates artificially 
important losses of precision (see \S\ref{ssec:stepbystep}). 
It is the so-called \emph{numerical instability} issue.

@@

Beyond sums and products (which can be treated as before), a basic 
operation involving polynomials is Euclidean division. Let us compute 
the differential of it. We start with two polynomials $A$ and $B$ with 
$B \neq 0$. The Euclidean division of $A$ by $B$ is written $A = BQ + R$ 
with $\deg R < \deg B$. Differentiating the above equality we find: 
$dA - dB \cdot Q = B \cdot dQ + dR$,
which implies that $dQ$ and $dR$ are respectively obtained as the 
quotient and the remainder of the Euclidean division of $dA - dB \cdot 
Q$ by $B$. This gives the differential. We note that the discussion 
above extends readily to convergent series (see also 
\cite{Caruso-Lubicz}).

@@

We now focus on the computation of Gr\"obner basis. Recall that any 
ideal $I \subset K[\XX]$ admits a unique reduced Gr\"obner basis. We can 
then consider the application mapping a family $(f_1, \ldots, f_s)$ of 
\emph{homogeneous}\footnote{This restriction will be convenient for us 
mainly because we are using the article \cite{Vaccon}, which deals only 
with homogeneous polynomials. It is however probably not essential.} 
polynomials of fixed degrees to the reduced Gr\"obner basis $(g_1, 
\ldots, g_t)$. It follows from Theorem 1.1 of \cite{Vaccon} that this 
application at $(f_1, \ldots, f_s)$ is continuous provided that:
\begin{itemize}
\item the sequence $(f_1, \ldots, f_s)$ is regular
\item for all $i \in \{1, \ldots, s\}$, the ideal generated by
$f_1, \ldots, f_i$ is wearly-$w$ where $w$ denotes the fixed monomial
order (\emph{cf} \cite{Vaccon} for more precisions).
\end{itemize}
A similar argument proves that it is actually differentiable at such
points. We are now going to compute the differential. For this we
remark that since the families $(f_1, \ldots, f_s)$ and $(g_1, \ldots,
g_t)$ generate the same ideal, we have a relation:
\begin{equation}
\label{eq:prodgrobner}
(g_1, \ldots, g_t) = (f_1, \ldots, f_s) \cdot A
\end{equation}
where $A$ is a $(s \times t)$ matrix with coefficients in $K[\XX]$.
Moreover following Vaccon's construction, we see that the entries of $A$ 
can be chosen in such a way that they define differentiable functions.
Differentiating Eq.~\eqref{eq:prodgrobner}, we get:
$$(d f_1, \ldots, d f_s) \cdot A + (f_1, \ldots, f_s) \cdot dA = 
(d g_1, \ldots, d g_t).$$
Finally, using that we are working with \emph{reduced} Gr\"obner
basis, the above equality implies that $d g_i$ is the remainder in
the division of the $i$-th coefficient of the product
$(d f_1, \ldots, d f_s) \cdot A$ by the family $(g_1, \ldots, g_t)$.

@@

\bibitem{Caruso-Lubicz}
  X.~Caruso, D.~Lubicz
  \emph{Linear Algebra over Zp[[u]] and related rings,} 
  to appear at LMS J. Comp. and Math.
